Stack Safety as a Security Property

### 2020-07-02 main.tex
Comments on the new running example

\begin{figure}[b]
  \begin{centering}
    \begin{minipage}{0.5\textwidth}
\begin{verbatim}
f{
  int x, y;
  ...
  y = g(0);
  ...
}

g(x){
  return (x + h(17));
}

h(x){
  return (x + 1);
}
\end{verbatim}
\end{minipage}
\end{centering}
\caption{High-level Program with Simple Nested Calls}
\label{fig:running-program}
\end{figure}

@@

Our running example consists of three functions $f$, $g$, and $h$ that
perform a nested sequence of calls. A high level C-like view of the
excerpt program can be found in Figure~\ref{fig:running-program}.
Suppose that the starting state for our running example is right
before $f$'s call to $g$. A rough example layout of the memory at that
point can be found in the top of Figure~\ref{fig:running-trace-b}:
\begin{itemize}
\item The low part of the memory (0-1000) is dedicated to instructions.
\item The stack starts at location 1000.
\item The stack of $f$ starts at 1040.
\item The location of $x$ is 1040 and stores $42$.
\item The location of $y$ is 1044 and stores $0$.
\end{itemize}
Directly below the layout, we summarize the parts of the register file
that are relevant to the example.

The rest of Figure~\ref{fig:running-trace-b} detail the execution of the
program, showing the machine state at call and return points, as well
as describing the full effect of each instruction in detail.
\leo{Unoptimized, etc.}

@@

@@

  \begin{array}{r|c|c|c|c}

@@

\begin{figure}
% After Return from H
% Memory Layout
\begin{center}
\MemoryLabel{1.7em}{1.8em}{0}
\MemoryLabel{21.2em}{1.8em}{1000}
\MemoryLabel{29.7em}{1.8em}{1040}
\MemoryLabel{49em}{1.8em}{$\downarrow$}
\MemoryLabel{48.8em}{3em}{SP}
\memory{6}{gray}[{\makebox[0pt]{Instructions}}]%
~$\cdots$
\memory{2}{gray}[{\makebox[0pt]{Stack}}]%
~$\cdots$
\memory{2}{green}[{\makebox[0pt]{$F_{priv}$}}]%
\memory{2}{yellow}[{\makebox[0pt]{arg-res}}]%
\memory{3}{red}[{\makebox[0pt]{ra-arg-res}}]%
\memory{3}{gray}
~$\cdots$
\MemoryLabel{-29.4em}{0.6em}{42}
\MemoryLabel{-27.2em}{0.6em}{0}
\MemoryLabel{-25.0em}{0.6em}{0}
\MemoryLabel{-22.8em}{0.6em}{0}
\MemoryLabel{-20.6em}{0.6em}{116}
\MemoryLabel{-18.4em}{0.6em}{17}
\MemoryLabel{-16.2em}{0.6em}{18}
\\
\end{center}
% Register File
\[
\begin{array}{cccccc}
  PC: 168 & r_0 : 0 & \rra : 168 & \rsp : 1068 & r_4 : 18 & r_5 : ? 
\end{array}
\]
% H - Body
\vspace*{0.2em}
\[
\begin{array}{r|c|c|c|c}
  \tracerow{20}{168}{\lw ~ r_4 ~ \rsp}{r_4 \leftarrow 18}{Load return value to $r_4$}
  \tracerow{21}{172}{\lw ~ r_5 ~ \rsp ~ -20}{r_5 \leftarrow 0}{Load the argument of $g$ to $r_5$}
  \tracerow{22}{176}{\add ~ r_4 ~ r_4 ~ r_5}{r_4 \leftarrow 18 + 0}{Add $r_4$ and $r_5$}
  \tracerow{23}{180}{\sw ~ \rsp ~ r_4 ~ -8}{[1056] \leftarrow 18}{Store $r_4$ as ther return of $g$}
  \tracerow{24}{184}{\addi ~ \rsp ~ \rsp ~ -8}{\rsp \leftarrow 1060}{Deallocate local state}
  \tracerow{25}{188}{\lw ~ \rra ~ \rsp ~ -4}{\rra \leftarrow 116}{Load return address}
  \tracerow{26}{192}{\addi ~ \rsp ~ \rsp ~ -4}{\rsp \leftarrow 1056}{Restore the stack pointer of $f$}
  \tracerow{27}{196}{\jalr ~ \rra ~ \rra ~ 0}{PC \leftarrow 116}{Return to $f$}
\end{array}
\]
% At Call to G
% Memory Layout
\begin{center}
\MemoryLabel{1.7em}{1.8em}{0}
\MemoryLabel{21.2em}{1.8em}{1000}
\MemoryLabel{29.7em}{1.8em}{1040}
\MemoryLabel{41em}{1.8em}{$\downarrow$}
\MemoryLabel{40.8em}{3em}{SP}
\memory{6}{gray}[{\makebox[0pt]{Instructions}}]%
~$\cdots$
\memory{2}{gray}[{\makebox[0pt]{Stack}}]%
~$\cdots$
\memory{2}{green}[{\makebox[0pt]{$F_{priv}$}}]%
\memory{2}{yellow}[{\makebox[0pt]{arg + result}}]%
\memory{6}{gray}
~$\cdots$
\MemoryLabel{-29.4em}{0.6em}{42}
\MemoryLabel{-27.2em}{0.6em}{0}
\MemoryLabel{-25.0em}{0.6em}{0}
\MemoryLabel{-22.8em}{0.6em}{18}
\\
\end{center}
% Register File
\[
\begin{array}{cccccc}
  PC: 116 & r_0 : 0 & \rra : 116 & \rsp : 1056 & r_4 : 18 & r_5 : 0
\end{array}
\]
\vspace*{0.2em}
\[
\begin{array}{r|c|c|c|c}
  \tracerow{28}{116}{\lw ~ r_4 ~ \rsp}{r_4 \leftarrow 18}{Load return value to $r_4$}
  \tracerow{29}{120}{\sw ~ \rsp ~ r_4 ~ -12}{[1048] \leftarrow 18}{Set $y$ to the return value}
  \tracerow{30}{124}{\addi ~ \rsp ~ \rsp ~ -8}{\rsp \leftarrow 1048}{Deallocate Call Frame}
\end{array}
\]
\vspace*{0.2em}  
% Ending Memory Layout
\begin{center}
\MemoryLabel{1.7em}{1.8em}{0}
\MemoryLabel{21.2em}{1.8em}{1000}
\MemoryLabel{29.7em}{1.8em}{1040}
\MemoryLabel{35.2em}{1.8em}{$\downarrow$}
\MemoryLabel{35em}{3em}{SP}
\memory{6}{gray}[{\makebox[0pt]{Instructions}}]%
~$\cdots$
\memory{2}{gray}[{\makebox[0pt]{Stack}}]%
~$\cdots$
\memory{2}{green}[{\makebox[0pt]{$F_{priv}$}}]%
\memory{8}{gray}
~$\cdots$
\MemoryLabel{-29.6em}{0.6em}{42}
\MemoryLabel{-27.4em}{0.6em}{0}
\\
\end{center}
% Starting Register File
\[
\begin{array}{cccccc}
  PC: 128 & r_0 : 0 & \rra : 116 & \rsp : 1048 & r_4 : 18 & r_5 : 0
\end{array}
\]
\caption{Detailed Execution Trace}
\label{fig:running-trace-b}
\end{figure}

@@

\begin{figure}
  % H - Body
  \[
  \begin{array}{r|c|c|c|c}
    \tracerow{1}{100}{\lw ~ r_0 ~ 40}{}{Load value at 40 to $r_0$}
    \tracerow{2}{104}{\sw ~ 42 ~ r_0}{}{Store that value to 42}
    \tracerow{3}{108}{\jalr ~ \rra ~ \rra ~ 0}{}{Return}
  \end{array}
  \]

@@

\begin{figure}
  % H - Body
  \[
  \begin{array}{r|c|c|c|c}
    \tracerow{1}{100}{\lw ~ r_0 ~ 40}{}{Load value at 40 to $r_0$}
    \tracerow{2}{104}{\sw ~ 42 ~ 0}{}{Store a constant value to 42}
    \tracerow{3}{108}{\lw ~ r_0 ~ 45}{}{Overwrite $r_0$ with accessible value}
    \tracerow{4}{112}{\jalr ~ \rra ~ \rra ~ 0}{}{Return}
  \end{array}
  \]

>>>>

\begin{figure}[b]
  \begin{centering}
    \begin{minipage}{.8\textwidth}
\begin{verbatim}
f = {                 g(x) = {                      h(x) = {
   int x, y;                    return (x + h(17));                  return (x + 1);
  ...                 }                             }
  y = g(0);
  ...
}
\end{verbatim}
\end{minipage}
\end{centering}
\caption{High-level Program with Simple Nested Calls \bcp{Rename parameters
    of g and h to z and w}}
\label{fig:running-program}
\end{figure}

@@

\bcp{
Comments on the example:
\begin{itemize}
\item We can simplify a bit by renaming $f$ to $\mathit{main}$ and
relocating its stack to location 1000.  And maybe move its instructions to
start at 0.  (We can comment that on a real RISC-V this is where interrupt
handlers go, but POPL readers will not care too much.)  We should streamline
the example as much as possible.
\item Do we need the ``...''s in $f$?
%
\item We should also emphasize, again, that there is no heap (if there were,
we might put it above the stack or between the instructions and the stack).
%
\item We will also need to be careful about fonts ($f$ vs {\texttt f}),
capitalization of register names (sp vs SP), etc. Maybe typeset instructions
in small caps?  Maybe typeset everything that is part of a concrete machine
in small caps??
%
\item The colors are helpful in making the big page of instruction traces easier
on the eye; maybe they would be even more helpful if
corresponding colors appeared elsewhere in the example (on the instructions
themselves, e.g.).  But we need to be careful to keep it readable -- maybe
they should be background colors, like you've done, rather than text colors.
\item The first column of numbers is redundant with the second column (PC).
Suggest deleting the first.
\item Asking readers to wade through all this is asking quite a bit!  We
need to make sure we are going to get sufficient value out of the example
later (I think we can, but we should really convince ourselves) to make it
worthwhile.
\item Use ``='' instead of ``:'' in the ``summaries'' below the pictures.
And put a bit more space between the components.
\item
\end{itemize}
}

Our running example consists of three functions $f$, $g$, and $h$ that
perform a nested sequence of calls. A high level C-like view of the
excerpt program can be found in Figure~\ref{fig:running-program}.
Suppose that the starting state for our running example is right
before $f$'s call to $g$. A sketch of the memory layout at that
point can be found in the top of Figure~\ref{fig:running-trace-b}:
\begin{itemize}
\item The low part of the memory (0-1000) is dedicated to instructions.
\item The stack starts at location 1000.
\item The stack of $f$ starts at 1040.
\item The location of $x$ is 1040 and stores $42$.
\item The location of $y$ is 1044 and stores $0$.
\end{itemize}
Directly below the layout, we summarize the parts of the register file
that are relevant to the example.

The rest of Figure~\ref{fig:running-trace-b} details the execution of the
program, showing the machine state at call and return points, as well
as describing the full effect of each instruction in detail.
\leo{Unoptimized, etc.}\bcp{??}

@@

\memory{2}{green}[{\makebox[0pt]{$F_{\textit{priv}}$}}]%

@@

  \begin{array}{l|l|l|l|l}

@@

\begin{figure}
% After Return from H
% Memory Layout
\begin{center}
\MemoryLabel{1.7em}{1.8em}{0}
\MemoryLabel{21.2em}{1.8em}{1000}
\MemoryLabel{29.7em}{1.8em}{1040}
\MemoryLabel{49em}{1.8em}{$\downarrow$}
\MemoryLabel{48.8em}{3em}{SP}
\memory{6}{gray}[{\makebox[0pt]{Instructions}}]%
~$\cdots$
\memory{2}{gray}[{\makebox[0pt]{Stack}}]%
~$\cdots$
\memory{2}{green}[{\makebox[0pt]{$F_{priv}$}}]%
\memory{2}{yellow}[{\makebox[0pt]{arg-res}}]%
\memory{3}{red}[{\makebox[0pt]{ra-arg-res}}]%
\memory{3}{gray}
~$\cdots$
\MemoryLabel{-29.4em}{0.6em}{42}
\MemoryLabel{-27.2em}{0.6em}{0}
\MemoryLabel{-25.0em}{0.6em}{0}
\MemoryLabel{-22.8em}{0.6em}{0}
\MemoryLabel{-20.6em}{0.6em}{116}
\MemoryLabel{-18.4em}{0.6em}{17}
\MemoryLabel{-16.2em}{0.6em}{18}
\\
\end{center}
% Register File
\[
\begin{array}{cccccc}
  PC: 168 & r_0 : 0 & \rra : 168 & \rsp : 1068 & r_4 : 18 & r_5 : ?
\end{array}
\]
% H - Body
\vspace*{0.2em}
\[
\begin{array}{l|l|l|l|l}
  \tracerow{20}{168}{\lw ~ r_4 ~ \rsp}{r_4 \leftarrow 18}{Load return value to $r_4$}
  \tracerow{21}{172}{\lw ~ r_5 ~ \rsp ~ -20}{r_5 \leftarrow 0}{Load the argument of $g$ to $r_5$}
  \tracerow{22}{176}{\add ~ r_4 ~ r_4 ~ r_5}{r_4 \leftarrow 18 + 0}{Add $r_4$ and $r_5$}
  \tracerow{23}{180}{\sw ~ \rsp ~ r_4 ~ -8}{[1056] \leftarrow 18}{Store $r_4$ as ther return of $g$}
  \tracerow{24}{184}{\addi ~ \rsp ~ \rsp ~ -8}{\rsp \leftarrow 1060}{Deallocate local state}
  \tracerow{25}{188}{\lw ~ \rra ~ \rsp ~ -4}{\rra \leftarrow 116}{Load return address}
  \tracerow{26}{192}{\addi ~ \rsp ~ \rsp ~ -4}{\rsp \leftarrow 1056}{Restore the stack pointer of $f$}
  \tracerow{27}{196}{\jalr ~ \rra ~ \rra ~ 0}{PC \leftarrow 116}{Return to $f$}
\end{array}
\]
% At Call to G
% Memory Layout
\begin{center}
\MemoryLabel{1.7em}{1.8em}{0}
\MemoryLabel{21.2em}{1.8em}{1000}
\MemoryLabel{29.7em}{1.8em}{1040}
\MemoryLabel{41em}{1.8em}{$\downarrow$}
\MemoryLabel{40.8em}{3em}{SP}
\memory{6}{gray}[{\makebox[0pt]{Instructions}}]%
~$\cdots$
\memory{2}{gray}[{\makebox[0pt]{Stack}}]%
~$\cdots$
\memory{2}{green}[{\makebox[0pt]{$F_{priv}$}}]%
\memory{2}{yellow}[{\makebox[0pt]{arg + result}}]%
\memory{6}{gray}
~$\cdots$
\MemoryLabel{-29.4em}{0.6em}{42}
\MemoryLabel{-27.2em}{0.6em}{0}
\MemoryLabel{-25.0em}{0.6em}{0}
\MemoryLabel{-22.8em}{0.6em}{18}
\\
\end{center}
% Register File
\[
\begin{array}{cccccc}
  PC: 116 & r_0 : 0 & \rra : 116 & \rsp : 1056 & r_4 : 18 & r_5 : 0
\end{array}
\]
\vspace*{0.2em}
\[
\begin{array}{l|l|l|l|l}
  \tracerow{28}{116}{\lw ~ r_4 ~ \rsp}{r_4 \leftarrow 18}{Load return value to $r_4$}
  \tracerow{29}{120}{\sw ~ \rsp ~ r_4 ~ -12}{[1048] \leftarrow 18}{Set $y$ to the return value}
  \tracerow{30}{124}{\addi ~ \rsp ~ \rsp ~ -8}{\rsp \leftarrow 1048}{Deallocate Call Frame}
\end{array}
\]
\vspace*{0.2em}
% Ending Memory Layout
\begin{center}
\MemoryLabel{1.7em}{1.8em}{0}
\MemoryLabel{21.2em}{1.8em}{1000}
\MemoryLabel{29.7em}{1.8em}{1040}
\MemoryLabel{35.2em}{1.8em}{$\downarrow$}
\MemoryLabel{35em}{3em}{SP}
\memory{6}{gray}[{\makebox[0pt]{Instructions}}]%
~$\cdots$
\memory{2}{gray}[{\makebox[0pt]{Stack}}]%
~$\cdots$
\memory{2}{green}[{\makebox[0pt]{$F_{priv}$}}]%
\memory{8}{gray}
~$\cdots$
\MemoryLabel{-29.6em}{0.6em}{42}
\MemoryLabel{-27.4em}{0.6em}{0}
\\
\end{center}
% Starting Register File
\[
\begin{array}{cccccc}
  PC: 128 & r_0 : 0 & \rra : 116 & \rsp : 1048 & r_4 : 18 & r_5 : 0
\end{array}
\]
\caption{Detailed Execution Trace}
\label{fig:running-trace-b}
\end{figure}

@@

\begin{figure}
  % H - Body
  \[
  \begin{array}{l|l|l|l|l}
    \tracerow{1}{100}{\lw ~ r_0 ~ 40}{}{Load value at 40 to $r_0$}
    \tracerow{2}{104}{\sw ~ 42 ~ r_0}{}{Store that value to 42}
    \tracerow{3}{108}{\jalr ~ \rra ~ \rra ~ 0}{}{Return}
  \end{array}
  \]

@@

\begin{figure}
  % H - Body
  \[
  \begin{array}{l|l|l|l|l}
    \tracerow{1}{100}{\lw ~ r_0 ~ 40}{}{Load value at 40 to $r_0$}
    \tracerow{2}{104}{\sw ~ 42 ~ 0}{}{Store a constant value to 42}
    \tracerow{3}{108}{\lw ~ r_0 ~ 45}{}{Overwrite $r_0$ with accessible value}
    \tracerow{4}{112}{\jalr ~ \rra ~ \rra ~ 0}{}{Return}
  \end{array}
  \]
